Leadership across cultures and contexts: an exploratory analysis

### 2020-05-09 leadership_across_cultures_contexts.Rmd
Spell check.

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over the intercept-only models (see Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying dimensions of leadership, one factor had a noticably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions. See Figure \@ref(fig:heatmapdrop1)A.

@@

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments*, *Control economics*, and *Control immigration*. Furthermore, evidence for leaders lacking coercive authority (our *No coercive authority* variable) was documented in over 35% of cultures (Figure \@ref(fig:varsupport)) and demonstrated variability by susbsitence strategy with hunter-gatherers providing relaitvely higher levels of evidence for leaders lacking coercive authority. See Figure S\@ref(fig:heatmapdrop1).

@@

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. Leaders function to resolve conflict, organize cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligent, and experienced and accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductivly, and socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). The composite feature variables mirrored these results, finding evidence in 85% of cultures or more, that leaders had social, reproductive, and material success, exhibited a number of competencies, organized and strategized group cooperation, and were prosocial. See Figure \@ref(fig:featuressupportplot).

@@

Leaders can be conceptualized as performing three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits charachteristic of leadership, leaders can be conceptualized as possessing some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural pressures, and recieving various benefits in exchange for their functions. The leadership "features" we derived by clustering leadership dimensions were represented across most cultures. See Figure \@ref(fig:featuressupportplot).

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories of leadership that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigous leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature might correspond, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2019].

The universal dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015] in which leaders recieve special status in exchange for the services they provide followers, but that model does not emphasize the concrete material benefits seen here. The absence of *High status* and *Aggressiveness* on this universal list corresponds to the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity and status leveling mechanisms.

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. But the computational services model emphasizes mating benefits, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality". See Figure \@ref(fig:varsupportcb).

Pawnee chiefs illustrate leadership involving several of the universal dimesions, including intelligence, conflict resolution, and generosity [@nq18-015]:

@@

Coercive authority is a dimension of leadership that is common in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authoiry among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controling economics, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies. See Figure \@ref(fig:elasticcoercive). This passage describing Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb). There was *systematic* variation in `r vary_n` leadership dimensions by group context, subsistence strategy, continental region, and/or leader sex, factors for which we had complete data for all text records. Interestingly, these systematically varying dimensions tended to be associated with one factor above the others. See Figure \@ref(fig:heatmapdrop1). The clearest pattern to emerge from these analyses is that, not too surprisingly, many dimensions of leadership are context-specfic: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group ledaership, and providing counsel and older age were associated with kin-group leadership. See Figure \@ref(fig:heatmapgroups).

@@

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of the nonindustiral cultures sampled here.

@@

Leadership and social influence expected to be associated with particular suites of characteristics and functions, which potentially covary by cultural or social context, personality, or other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @bass_handbook_1990]. Our systematic cross-cultural evidence supports many aspects of influential theories. Much of the evolutionary theoretical literature on leadership, however, has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social heirarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_j.k_dominance_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance and prestige as distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions across cultures and contexts, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, impliment strategic actions, and provide prosocial services to the group, all while conforming to cultural pressures. Compartamentalizing this range of diversity into two distinct social stratiges, although theoretically and empirically convenient, likely misses much of the ethnographic reality of human leadership dynamics.

>>>>

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over the intercept-only models (see Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying dimensions of leadership, one factor had a noticeably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions. See Figure \@ref(fig:heatmapdrop1)A.

@@

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments*, *Control economics*, and *Control immigration*. Furthermore, evidence for leaders lacking coercive authority (our *No coercive authority* variable) was documented in over 35% of cultures (Figure \@ref(fig:varsupport)) and demonstrated variability by subsistence strategy with hunter-gatherers providing relatively higher levels of evidence for leaders lacking coercive authority. See Figure S\@ref(fig:heatmapdrop1).

@@

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. Leaders function to resolve conflict, organize cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligent, and experienced and accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, and socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). The composite feature variables mirrored these results, finding evidence in 85% of cultures or more, that leaders had social, reproductive, and material success, exhibited a number of competencies, organized and strategized group cooperation, and were prosocial. See Figure \@ref(fig:featuressupportplot).

@@

Leaders can be conceptualized as performing three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits characteristic of leadership, leaders can be conceptualized as possessing some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural pressures, and receiving various benefits in exchange for their functions. The leadership "features" we derived by clustering leadership dimensions were represented across most cultures. See Figure \@ref(fig:featuressupportplot).

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories of leadership that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature might correspond, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2019].

The universal dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015] in which leaders receive special status in exchange for the services they provide followers, but that model does not emphasize the concrete material benefits seen here. The absence of *High status* and *Aggressiveness* on this universal list corresponds to the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity and status leveling mechanisms.

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. But the computational services model emphasizes mating benefits, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality". See Figure \@ref(fig:varsupportcb).

Pawnee chiefs illustrate leadership involving several of the universal dimensions, including intelligence, conflict resolution, and generosity [@nq18-015]:

@@

Coercive authority is a dimension of leadership that is common in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authority among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controlling economics, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies. See Figure \@ref(fig:elasticcoercive). This passage describing Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb). There was *systematic* variation in `r vary_n` leadership dimensions by group context, subsistence strategy, continental region, and/or leader sex, factors for which we had complete data for all text records. Interestingly, these systematically varying dimensions tended to be associated with one factor above the others. See Figure \@ref(fig:heatmapdrop1). The clearest pattern to emerge from these analyses is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership. See Figure \@ref(fig:heatmapgroups).

@@

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of the nonindustrial cultures sampled here.

@@

Leadership and social influence expected to be associated with particular suites of characteristics and functions, which potentially covary by cultural or social context, personality, or other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @bass_handbook_1990]. Our systematic cross-cultural evidence supports many aspects of influential theories. Much of the evolutionary theoretical literature on leadership, however, has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_j.k_dominance_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance and prestige as distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions across cultures and contexts, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural pressures. Compartmentalizing this range of diversity into two distinct social strategies, although theoretically and empirically convenient, likely misses much of the ethnographic reality of human leadership dynamics.
